Component response rate variation drives stability in large complex systems

### 2018-06-01 notebook/ms.Rmd
Minor edits, particularly emphasising the genetic algorithm results

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA alexander.duthie@stir.ac.uk
output:
  html_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

Rows in $M$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a real scalar value $\gamma_{i}$[@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $M$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $M$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $\gamma_{i}$ to generate $M$. When $\gamma_{i} = 1$, eigenvalues of $M$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (grey circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$) and all $\Re({\lambda}) < 0$. The latter cluster has real components that have shifted to the right, but has a smaller radius. Overall, for 1 million randomly assembled $M$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,1])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from $2-32$ (see Supplementary Information for different values of $\sigma$ and $C$). One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$ (note that under these conditions, $\sigma\sqrt{SC} = 1$ given $S = 25$ when $\gamma = 1$). I found that the number of stable random systems was consistently higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$, and the maximum observed $S$ for which a system was stable was $31$ given $Var(\gamma)$ versus $27$ given $\gamma = 1$ (see Supplementary Information for full results). This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise that $Var(\gamma)$ is not stabilising per se; that is, adding variation in $\gamma$ to a particular system $M$ does not necessarily increase the probability that the system will be stable (see Supplementary Information). Rather, systems that are observed to be stable are more likely to vary in $\gamma$, and for this $Var(\gamma)$ to be critical to their stability. This is caused by the shift in the distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes result in all $\Re(\lambda) < 0$ but might also increase $\Re(\lambda)$ values. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a simple genetic algorithm (the space of possible $\gamma$ values was too large to searh exhaustively[@Hamblin2013]; see Supplementary Information). For each of 10000 random $M$, the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis of each set of $\gamma$ values was performed, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a converge criteria of minimum fitness increase between generations was satisfied. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were frequently found that resulted in stable systems (the highest at $S = 39$), suggesting that varying component response time might by itself be sufficient to stabilise complex systems.

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina2012]. These include systems in which all interactions (i.e., all off-diagonal elements of $M$) are negative (e.g., competitive networks), positive (e.g., mutualist networks), or $i$ and $j$ pairs have opposing signs (e.g., predator-prey networks). In general, competitive and mutualist networks tend to be destabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist communities does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate[@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA alexander.duthie@stir.ac.uk
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

Rows in $M$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a real scalar value $\gamma_{i}$[@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $M$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $M$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn randomly from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $\gamma_{i}$ to generate $M$. When $\gamma_{i} = 1$, eigenvalues of $M$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (grey circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$) and all $\Re({\lambda}) < 0$. The latter cluster has real components that have shifted to the right, but has a smaller radius. Overall, for 1 million randomly assembled $M$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,1])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from 2-32 (see Supplementary Information for different values of $\sigma$ and $C$). One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$ (note that under these conditions, $\sigma\sqrt{SC} = 1$ given $S = 25$ when $\gamma = 1$). I found that the number of stable random systems was consistently higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$, and the maximum observed $S$ for which a system was stable was $31$ given $Var(\gamma)$ versus $27$ given $\gamma = 1$ (see Supplementary Information for full results). This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

Some caution is needed when interpreting these results. First, I emphasise that $Var(\gamma)$ is not stabilising per se; that is, adding variation in $\gamma$ to a particular system $M$ does not necessarily increase the probability that the system will be stable (see Supplementary Information). Rather, systems that are observed to be stable are more likely to vary in $\gamma$, and for this $Var(\gamma)$ to be critical to their stability. This is caused by the shift in the distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes result in all $\Re(\lambda) < 0$ but might also increase $\Re(\lambda)$ values. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a simple genetic algorithm (the space of possible $\gamma$ values was too large to search exhaustively[@Hamblin2013]; see Supplementary Information). For each of 10000 random $M$, the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis of each set of $\gamma$ values was performed, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a converge criteria of minimum fitness increase between generations was satisfied. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were orders of magnitude higher than when $\gamma = 1$ (see Supporting Information). Hence, managing the response rates of system components in a targetted way can potentially facilitate the stabilisation of complex systems.

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina2012]. These include systems in which all interactions are negative (e.g., competitive networks), positive (e.g., mutualist networks), or $i$ and $j$ pairs have opposing signs (e.g., predator-prey networks). In general, competitive and mutualist networks tend to be destabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist networks does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate[@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

### 2018-06-01 notebook/ms.tex
Minor edits, particularly emphasising the genetic algorithm results

Rows in \(M\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a real scalar
value
\(\gamma_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{14}}.
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(M\) where \(\gamma_{i} = 1\) for all \(i\) in \(S\) to
the same \(M\) when half of \(\gamma_{i} = 1.95\) and half of
\(\gamma_{i} = 0.05\). This models one system in which \(\gamma\) is
invariant and one in which \(\gamma\) varies, but systems are otherwise
identical (note \(E[\gamma_{i}] = 1\) in both cases). I assume
\(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal elements are
set to \(-1\) and non-zero off-diagonal elements are drawn from
\(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied by
\(\gamma_{i}\) to generate \(M\). When \(\gamma_{i} = 1\), eigenvalues
of \(M\) are distributed uniformly within a circle centred at
(\(-1, 0\)) with a radius of 1.265 (Fig. 1a). Hence, the real components
of eigenvalues are highly unlikely to all be negative when all
\(\gamma_{i} = 1\). But when \(\gamma_{i}\) values are separated into
two groups, eigenvalues are no longer uniformly distributed (Fig. 1b).
Instead, two distinct clusters of eigenvalues appear (grey circles in
Fig. 1b), one centred at (\(-1.95, 0\)) and the other centred at
(\(-0.05, 0\)). The former has a large radius, but the real components
have shifted to the left (in comparison to when \(\gamma = 1\)) and all
\(\Re({\lambda}) < 0\). The latter cluster has real components that have
shifted to the right, but has a smaller radius. Overall, for 1 million
randomly assembled \(M\), this division between slow and fast component
response rates results in more stable systems: 1 stable given
\(\gamma = 1\) versus 32 stable given \(\gamma = \{1.95, 0.5\}\).

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.551 and 0.553 for a and b, respectively.

@@

To investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from \(2-32\) (see Supplementary Information for different
values of \(\sigma\) and \(C\)). One million \(M\) were simulated for
each \(S\), and the stability of \(M\) was assessed given \(\gamma = 1\)
versus \(\gamma \sim \mathcal{U}(0, 2)\) (note that under these
conditions, \(\sigma\sqrt{SC} = 1\) given \(S = 25\) when
\(\gamma = 1\)). I found that the number of stable random systems was
consistently higher given \(Var(\gamma)\) than when \(\gamma = 1\) (Fig.
3), and that the difference between the probabilities of observing a
stable system increased with an increase in \(S\); i.e., the potential
for \(Var(\gamma)\) to drive stability increased with system complexity.
For the highest values of \(S\), nearly all systems that were stable
given \(Var(\gamma)\) would not have been stable given \(\gamma = 1\),
and the maximum observed \(S\) for which a system was stable was \(31\)
given \(Var(\gamma)\) versus \(27\) given \(\gamma = 1\) (see
Supplementary Information for full results). This suggests that the
stability of large systems might be dependent upon variation in the
response rate of their individual components, meaning that factors such
as generation time (in ecological networks), transaction speed (in
economic networks), or communication speed (in social networks) needs to
be considered when investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise
that \(Var(\gamma)\) is not stabilising per se; that is, adding
variation in \(\gamma\) to a particular system \(M\) does not
necessarily increase the probability that the system will be stable (see
Supplementary Information). Rather, systems that are observed to be
stable are more likely to vary in \(\gamma\), and for this
\(Var(\gamma)\) to be critical to their stability. This is caused by the
shift in the distribution of eigenvalues that occurs by introducing
\(Var(\gamma)\) (Fig. 1b, 2b), which can sometimes result in all
\(\Re(\lambda) < 0\) but might also increase \(\Re(\lambda)\) values. To
further investigate the potential of \(Var(\gamma)\) to be stabilising,
I used a simple genetic algorithm (the space of possible \(\gamma\)
values was too large to searh
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{18}};
see Supplementary Information). For each of 10000 random \(M\), the
genetic algorithm initialised 1000 different sets of
\(\gamma \sim \mathcal{U}(0, 2)\) values of size \(S\). Eigenanalysis of
each set of \(\gamma\) values was performed, and the 20 sets with the
lowest \(\max\left(\Re(\lambda)\right)\) each produced 50 offspring with
subsequent mutation and crossover between the resulting new population
of 1000 \(\gamma\) sets. The genetic algorithm terminated if a stable
\(M\) was found, 20 generations occurred, or a converge criteria of
minimum fitness increase between generations was satisfied. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were frequently
found that resulted in stable systems (the highest at \(S = 39\)),
suggesting that varying component response time might by itself be
sufficient to stabilise complex systems.

I have focused broadly on random complex systems, but it is also
worthwhile to consider more restricted interactions such as those of
specific ecological
networks\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. These
include systems in which all interactions (i.e., all off-diagonal
elements of \(M\)) are negative (e.g., competitive networks), positive
(e.g., mutualist networks), or \(i\) and \(j\) pairs have opposing signs
(e.g., predator-prey networks). In general, competitive and mutualist
networks tend to be destabilising, and predator-prey network tend to be
stabilising\textsuperscript{\protect\hyperlink{ref-Allesina2011}{19}}.
When \(Var(\gamma)\) is applied to each, the proportion of stable
competitive and predator-prey networks increases, but the proportion of
stable mutualist communities does not (see Supplementary Information).
Additionally, when each component of \(M\) is interpreted as a unique
species and given a random intrinsic growth
rate\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}},
feasibility is not increased by \(Var(\gamma)\), suggesting that
variation in species generation time might be unlikely to drive
stability in purely multi-species networks (see Supplementary
Information).

>>>>

Rows in \(M\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a real scalar
value
\(\gamma_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{14}}.
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(M\) where \(\gamma_{i} = 1\) for all \(i\) in \(S\) to
the same \(M\) when half of \(\gamma_{i} = 1.95\) and half of
\(\gamma_{i} = 0.05\). This models one system in which \(\gamma\) is
invariant and one in which \(\gamma\) varies, but systems are otherwise
identical (note \(E[\gamma_{i}] = 1\) in both cases). I assume
\(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal elements are
set to \(-1\) and non-zero off-diagonal elements are drawn randomly from
\(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied by
\(\gamma_{i}\) to generate \(M\). When \(\gamma_{i} = 1\), eigenvalues
of \(M\) are distributed uniformly within a circle centred at
(\(-1, 0\)) with a radius of 1.265 (Fig. 1a). Hence, the real components
of eigenvalues are highly unlikely to all be negative when all
\(\gamma_{i} = 1\). But when \(\gamma_{i}\) values are separated into
two groups, eigenvalues are no longer uniformly distributed (Fig. 1b).
Instead, two distinct clusters of eigenvalues appear (grey circles in
Fig. 1b), one centred at (\(-1.95, 0\)) and the other centred at
(\(-0.05, 0\)). The former has a large radius, but the real components
have shifted to the left (in comparison to when \(\gamma = 1\)) and all
\(\Re({\lambda}) < 0\). The latter cluster has real components that have
shifted to the right, but has a smaller radius. Overall, for 1 million
randomly assembled \(M\), this division between slow and fast component
response rates results in more stable systems: 1 stable given
\(\gamma = 1\) versus 32 stable given \(\gamma = \{1.95, 0.5\}\).

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.549 and 0.557 for a and b, respectively.

@@

To investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from 2-32 (see Supplementary Information for different
values of \(\sigma\) and \(C\)). One million \(M\) were simulated for
each \(S\), and the stability of \(M\) was assessed given \(\gamma = 1\)
versus \(\gamma \sim \mathcal{U}(0, 2)\) (note that under these
conditions, \(\sigma\sqrt{SC} = 1\) given \(S = 25\) when
\(\gamma = 1\)). I found that the number of stable random systems was
consistently higher given \(Var(\gamma)\) than when \(\gamma = 1\) (Fig.
3), and that the difference between the probabilities of observing a
stable system increased with an increase in \(S\); i.e., the potential
for \(Var(\gamma)\) to drive stability increased with system complexity.
For the highest values of \(S\), nearly all systems that were stable
given \(Var(\gamma)\) would not have been stable given \(\gamma = 1\),
and the maximum observed \(S\) for which a system was stable was \(31\)
given \(Var(\gamma)\) versus \(27\) given \(\gamma = 1\) (see
Supplementary Information for full results). This suggests that the
stability of large systems might be dependent upon variation in the
response rate of their individual components, meaning that factors such
as generation time (in ecological networks), transaction speed (in
economic networks), or communication speed (in social networks) needs to
be considered when investigating the stability of complex systems.

Some caution is needed when interpreting these results. First, I
emphasise that \(Var(\gamma)\) is not stabilising per se; that is,
adding variation in \(\gamma\) to a particular system \(M\) does not
necessarily increase the probability that the system will be stable (see
Supplementary Information). Rather, systems that are observed to be
stable are more likely to vary in \(\gamma\), and for this
\(Var(\gamma)\) to be critical to their stability. This is caused by the
shift in the distribution of eigenvalues that occurs by introducing
\(Var(\gamma)\) (Fig. 1b, 2b), which can sometimes result in all
\(\Re(\lambda) < 0\) but might also increase \(\Re(\lambda)\) values. To
further investigate the potential of \(Var(\gamma)\) to be stabilising,
I used a simple genetic algorithm (the space of possible \(\gamma\)
values was too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{18}};
see Supplementary Information). For each of 10000 random \(M\), the
genetic algorithm initialised 1000 different sets of
\(\gamma \sim \mathcal{U}(0, 2)\) values of size \(S\). Eigenanalysis of
each set of \(\gamma\) values was performed, and the 20 sets with the
lowest \(\max\left(\Re(\lambda)\right)\) each produced 50 offspring with
subsequent mutation and crossover between the resulting new population
of 1000 \(\gamma\) sets. The genetic algorithm terminated if a stable
\(M\) was found, 20 generations occurred, or a converge criteria of
minimum fitness increase between generations was satisfied. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were orders of
magnitude higher than when \(\gamma = 1\) (see Supporting Information).
Hence, managing the response rates of system components in a targetted
way can potentially facilitate the stabilisation of complex systems.

I have focused broadly on random complex systems, but it is also
worthwhile to consider more restricted interactions such as those of
specific ecological
networks\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. These
include systems in which all interactions are negative (e.g.,
competitive networks), positive (e.g., mutualist networks), or \(i\) and
\(j\) pairs have opposing signs (e.g., predator-prey networks). In
general, competitive and mutualist networks tend to be destabilising,
and predator-prey network tend to be
stabilising\textsuperscript{\protect\hyperlink{ref-Allesina2011}{19}}.
When \(Var(\gamma)\) is applied to each, the proportion of stable
competitive and predator-prey networks increases, but the proportion of
stable mutualist networks does not (see Supplementary Information).
Additionally, when each component of \(M\) is interpreted as a unique
species and given a random intrinsic growth
rate\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}},
feasibility is not increased by \(Var(\gamma)\), suggesting that
variation in species generation time might be unlikely to drive
stability in purely multi-species networks (see Supplementary
Information).
